Viscous Transport in Eroding Porous Media

### 2019-09-23 docs/transport/erosion_transport.tex
went through intro

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Porous media flow plays an important role in many environmental and
industrial applications.  Depending on the application, length scales
can vary from $10^{-6}\mathrm{m}$ to
$10^{-1}\mathrm{m}$~\cite{mil-chr-imh-mcb-ped1998}, velocity scales can
be as small as $10^{-1}\mathrm{m/day}$~\cite{kut-scr-dav-ham1995}, and
the scales of a single porous media can range over several orders of
magnitude. Numerical methods that accurately address all of these scales
are essential to develop meaningful constitutive relationships between
the microscopic and macroscopic scales~\cite{mil-chr-imh-mcb-ped1998},
to characterize dispersion~\cite{saf1959}, and to quantify
mixing~\cite{leb-den-dav-bol-car-dec-bou2011, den-leb-eng-bij2011}.
Flow in porous media is further complicated when there is a two-way
coupling between the flow and the geometry. For example, the grains of a
porous media evolve in applications involving
melting~\cite{bec-vis1998}, dissolution~\cite{kan-zha-che-he2002},
deposition~\cite{joh-eli1995}, and biofilm
growth~\cite{tan-val-wer2015}.  We focus on erosion, a fluid-mechanical
process that is prevalent in many geophysical and hydrological
applications.

Examples of coarse-grained models for porous media flow include
pressure-saturation-permeability
relationships~\cite{mil-chr-imh-mcb-ped1998}, permeability-porosity
relationships~\cite{dar-mcc1998, car1937, won-kop-tom1984}, first
passage time distributions~\cite{ber-sch-sil2000, hym-den-hag-kan2019,
cve-che-wen1996}, tortuosities~\cite{hak-com-den2019, mat-kha-koz2008,
dud-koz-mat2011, kop-kat-tim1996}, concentration
distributions~\cite{ica-den2019, bel-sal-rin1992}, flow
distributions~\cite{ali-par-wei-bre2017}, geometry
connectivity~\cite{knu-car2005, wes-blo-gra2001}, pore
distributions~\cite{ali-par-wei-bre2017}, anomalous
dispersion~\cite{dea-qua-bir-jua2018, den-cor-sch-ber2004,
sie-ili-pri-riv-gua2019}, and more.  Some of these works consider
heterogeneous porous geometries, but to the best of our knowledge, a
detailed study of transport in eroded geometries has not been performed.
This paper characterizes transport in eroded geometries by computing and
analyzing a variety of coarse-grained variables.

When a porous geometry erodes, certain qualitative characteristics are
unveiled that affect transport through the geometry. For example, an
eroded geometry often contains channels of high porosity that connect
the inlet to the outlet, and this results in velocities that vary over
several orders of magnitude~\cite{all-hea-lab-rei2002}.  Moreover,
channelized geometries are heterogeneous and anisotropic, and this
affects the transfer of quantities such as
contaminants~\cite{cve-che-wen1996, dag1987, kon-bre1978} and
heat~\cite{nil-sto1990, ree-sto1995}.  To characterize the transport in
eroded geometries, we must complete three main tasks.  The first two are
forming high-fidelity numerical solutions of the fluid equations, and
eroding the grains according to an erosion model. These two tasks are
performed using our previous work~\cite{qua-moo2018}, but we improve our
numerical methods so that simulations of more realistic dense
suspensions of eroding grains are possible
(Figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  Finally,
in this paper, we simulate and characterize transport through eroded
porous geometries.

@@

Because of the length and velocity scales in hydrological applications
such as groundwater flow~\cite{bea1972}, we model the hydrodynamics with
the two-dimensional incompressible Stokes equations. A variety of
numerical methods are available to solve the incompressible Stokes
equations including lattice-gas or lattice
Boltzmann~\cite{kop-kat-tim1996, dar-mcc1998}, finite
volumes~\cite{suo-liu-gan2019, den-ica-hid2018,
sie-ili-pri-riv-gua2019}, finite differences~\cite{leb-ded-dav-bou2007,
knu-car2005}, and cellular automata~\cite{rot1988}.  However, to
accurately simulate erosion and transport, we require a high-order
method that is able to resolve geometries with arbitrarily shaped and
nearly-touching grains.  Therefore, we solve the incompressible Stokes
equations with a boundary integral equation (BIE).  To couple erosion
with the fluid, the grains are eroded at a rate that is proportional to
the magnitude of the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, par-izu2000}.  Since the vorticity and the
shear stress are equivalent on the grains in the zero Reynolds number
limit, we also compute the vorticity in the fluid bulk
(Figure~\ref{fig:Eroding50vort}).

To characterize the transport, particle trajectories need to be
computed. Depending on the application, microscale transport can be
modelled as pure advection~\cite{dea-qua-bir-jua2018,
leb-ded-dav-bou2007, cve-che-wen1996, puy-gou-den2019},
advection-diffusion~\cite{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\cite{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume that the trajectories are governed purely by
advection (ie.~no diffusion).  Therefore, a trajectory $\ss(t)$
initialized at $\ss_0$ is governed by
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity, and the trajectories are simply the
streamlines. Since there is no stiff diffusive term
in~\eqref{eqn:tracers}, we use a high-order explicit Runge-Kutta time
stepping method. Other methods to compute streamlines include a
continuous time random walk~\cite{mon-wei1965, den-cor-sch-ber2004,
leb-den-car2008, ber-cor-den-sch2006}.

\begin{figure}[H]
\begin{center}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b1}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b80}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b160}\\
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b240}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b320}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b400}
\caption{\label{fig:Eroding50vort} The same six snapshots from
Figure~\ref{fig:Eroding50vel}.  The color is the vorticity of the fluid.
Since the rate of erosion is equivalent to the magnitude of the
vorticity, erosion is fastest in the yellow and blue regions and slowest
in the green regions.}
\end{center}
\end{figure}

To be able to compute stable simulations of erosion and accurate
streamline trajectories, we use high-order methods in both space and
time. The erosion rate is singular at stagnation points, so we apply a
mild regularization and smoothing terms to eliminate numerical
instabilities.  As described in our previous work~\cite{qua-moo2018}, we
apply a stable second-order Runge-Kutta method is applied to the {\thL}
coordinates~\cite{hou-low-she1994} of the eroding grains.  In these
coordinates, the stiff term is linear and treated implicitly, so we are
able to take large time step sizes. In our previous work, after
discretizing the BIE, the trapezoid rule is used to approximate the
integrals. The accuracy of the trapezoid rule is sufficiently accurate
if all the bodies are sufficiently separated~\cite{tre-wei2014}, but
nearly-singular integrals must be computed when two grains are
nearly-touching or when evaluating the velocity or vorticity near an
eroding grain.  One of the earliest works on near-singular integration
was developed by Baker and Shelley~\cite{bak-she1986}, and in recent
years, many other near-singular integration schemes have been
developed~\cite{oja-tor2015, kli-tor2018, hel-oja2008a, bea-yin-wil2016,
bea-lai2001, klo-bar-gre-one2013}.  In this work, we use a Barycentric
quadrature method~\cite{bar2014, bar-wu-vee2015} since it is a
non-intrusive modification of the trapezoid rule, and the error is
guaranteed to be uniformly bounded.  We extend the original quadrature
method so that it can be used to compute the gradient of the velocity
that is required to compute the shear stress and the fluid vorticity. In
addition, the quadrature method is used to accurately compute streamline
trajectories that pass through the very slow regions close to the
eroding grains.

Once streamlines in an eroded geometry are computed, we characterize the
transport by computing and analyzing three different metrics: the
tortuosity, the anomalous dispersion, and the pore size distribution.
The local tortuosity of a streamline connecting the inlet to the outlet
is the streamline's length normalized by the distance between the inlet
and the outlet.  In porous media, depending on several factors including
the porosity, the local tortuosity can be greater than
1.5~\cite{kop-kat-tim1996, mat-kha-koz2008}, or even
2~\cite{dud-koz-mat2011}.  The tortuosity of a geometry is defined by
averaging the local tortuosity over all streamlines that are initialized
at the inlet, and the geometry's tortuosity can be used to characterize
average particle motions~\cite{hak-com-den2019}.  To characterize
spreading, the fluid dispersion is computed as the variance of the
streamline lengths. In porous media, this spreading can often be
characterized with a super-diffusive anomalous dispersion
rate~\cite{kan-dea-nun-bij-blu-jua2014, cus-hu-den1995,
dea-leb-den-tar-bol-dav2013}. An accurate solution near the grains is
crucial since the anomalous dispersion is caused by streamlines spending
time in both the high and low velocity regimes~\cite{ber-sch2001}.
Finally, the third metric, pore size distribution, is useful to quantify
velocity distributions~\cite{ali-par-wei-bre2017, dea-qua-bir-jua2018},
channelization~\cite{sie-ili-pri-riv-gua2019},
connectivity~\cite{knu-car2005, wes-blo-gra2001}, and to develop network
models~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
Our first contribution is extending the work of Barnett et
al.~\cite{bar-wu-vee2015} so that high-order accurate derivatives of
solutions of the Stokes equations can be computed.  Using this new
quadrature method, we accurately compute erosion rates and stably
simulate erosion.  Moreover, we use the quadrature method to visualize
the vorticity in the fluid bulk.  The vorticity is convenient to
visualize since, in the zero Reynolds number limit, it is equivalent to
the shear stress on the eroding bodies, and therefore its magnitude is
the erosion rate.

Our second contribution is computing and analyzing trajectories in the
eroded geometries.  These trajectories are used to characterize the
transport by computing local tortuosities of the trajectories, the
tortuosity of the eroded geometry, and the anomalous dispersion rates in
the geometry.  By using the same quadrature rule and a fourth-order
Runge-Kutta method, we are able to accurately compute trajectories that
spend large amounts of time in the slow velocity regions near the
eroding bodies, and computing these trajectories reliably is essential
to characterize the flow~\cite{dea-leb-den-tar-bol-dav2013}.  We compare
our tortuosity and dispersion rates with models available in the
literature.  Finally, we analyze the effect of erosion on the pore size
distribution.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations}
As in our previous work~\cite{qua-moo2018}, we consider two-dimensional
immobile eroding grains. However, this new work does not have a limitation on
the proximity of the bodies. Compared to our previous work, we are able to
simulate much denser suspensions with a comparable number of
discretization points and without upsampling.  While we do investigate
transport in this work, we only consider passive particles that are only
advected by the flow.  To model other physical processes, such as
contaminant transport, we plan to simulate a finite Peclet number
advection-diffusion equation in eroded geometries.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Porous media flow plays an important role in many environmental and
industrial applications.  Depending on the application, length scales
can vary from $10^{-6}\mathrm{m}$ to
$10^{-1}\mathrm{m}$~\cite{mil-chr-imh-mcb-ped1998}, velocity scales can
be as small as $10^{-1}\mathrm{m/day}$~\cite{kut-scr-dav-ham1995}, and
the scales of a single porous media can range over several orders of
magnitude. Numerical methods that accurately address all of these scales
are essential to develop meaningful constitutive relationships between
the microscopic and macroscopic scales~\cite{mil-chr-imh-mcb-ped1998},
to characterize dispersion~\cite{saf1959}, and to quantify
mixing~\cite{leb-den-dav-bol-car-dec-bou2011, den-leb-eng-bij2011}.
Examples of coarse-grained models for porous media flow include
pressure-saturation-permeability
relationships~\cite{mil-chr-imh-mcb-ped1998}, permeability-porosity
relationships~\cite{dar-mcc1998, car1937, won-kop-tom1984}, first
passage time distributions~\cite{ber-sch-sil2000, hym-den-hag-kan2019,
cve-che-wen1996}, tortuosities~\cite{hak-com-den2019, mat-kha-koz2008,
dud-koz-mat2011, kop-kat-tim1996}, concentration
distributions~\cite{ica-den2019, bel-sal-rin1992}, flow
distributions~\cite{ali-par-wei-bre2017}, geometry
connectivity~\cite{knu-car2005, wes-blo-gra2001}, pore
distributions~\cite{ali-par-wei-bre2017}, anomalous
dispersion~\cite{dea-qua-bir-jua2018, den-cor-sch-ber2004,
sie-ili-pri-riv-gua2019}, and more.

Flow in porous media is further complicated when there is a two-way
coupling between the flow and the geometry. For example, the grains of a
porous media evolve in time in applications involving
melting~\cite{bec-vis1998}, dissolution~\cite{kan-zha-che-he2002},
deposition~\cite{joh-eli1995}, and biofilm
growth~\cite{tan-val-wer2015}.  We focus on erosion, a fluid-mechanical
process that is prevalent in many geophysical and hydrological
applications.  When a porous geometry erodes, certain qualitative
characteristics are unveiled that affect transport through the geometry.
For example, an eroded geometry often contains channels of high porosity
that connect the inlet to the outlet, and this results in velocities
that vary over several orders of magnitude~\cite{all-hea-lab-rei2002}.
Moreover, channelized geometries are heterogeneous and anisotropic, and
this affects the transport of quantities such as
contaminants~\cite{cve-che-wen1996, dag1987, kon-bre1978} and
heat~\cite{nil-sto1990, ree-sto1995}.  To characterize transport in
eroded geometries, we complete two undertakings: forming realistic
eroded porous geometries and characterizing transport through these
geometries.  The modelling effort and numerical methods used to simulate
erosion are largely based on our previous work~\cite{qua-moo2018}.
However, we improve our numerical methods so that simulations of more
realistic dense suspensions of eroding grains are possible
(Figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  To
characterize transport in eroded porous geometries, we apply statistical
methods to develop coarse-grained variables.

@@

Because of the length and velocity scales in hydrological applications
such as groundwater flow~\cite{bea1972}, we model the hydrodynamics with
the two-dimensional incompressible Stokes equations.  To couple erosion
with the fluid, the grains are eroded at a rate that is proportional to
the magnitude of the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, par-izu2000}.  Since the vorticity and the
shear stress are equivalent on the grains in the zero Reynolds number
limit, we also compute the vorticity in the fluid bulk
(Figure~\ref{fig:Eroding50vort}).  A variety of numerical methods for
the fluid solver are available including lattice-gas or lattice
Boltzmann~\cite{kop-kat-tim1996, dar-mcc1998}, finite
volumes~\cite{suo-liu-gan2019, den-ica-hid2018,
sie-ili-pri-riv-gua2019}, finite differences~\cite{leb-ded-dav-bou2007,
knu-car2005}, and cellular automata~\cite{rot1988}.  However, simulating
erosion requires a method that is high-order, can resolve
nearly-touching arbitrarily shaped grains, and can accurately compute
derivatives of the velocity field.  Since the governing equations are
linear and homogeneous, we address these challenges by converting the
fluid equations to a boundary integral equation (BIE).

To compute stable simulations of erosion, we use high-order methods in
both space and time. The time integration method is unchanged from our
previous work~\cite{qua-moo2018}.  We apply a mild regularization and a
smoothing term to eliminate numerical instabilities caused by changes in
sign of the shear stress, and we use a stable second-order Runge-Kutta
method applied to the {\thL} coordinates~\cite{hou-low-she1994} of the
eroding grains.  In this work, we do introduce a new quadrature method
to resolve dense suspensions of eroding grains.  The accuracy of the
trapezoid rule, which we used in our previous work, is sufficient for
bodies that are sufficiently separated~\cite{tre-wei2014}. However, it
is inaccurate when two grains are nearly-touching.  One of the earliest
works that addresses quadrature methods for nearly-singular integrands
was developed by Baker and Shelley~\cite{bak-she1986}, and in recent
years, many other near-singular integration schemes have been
developed~\cite{oja-tor2015, kli-tor2018, hel-oja2008a, bea-yin-wil2016,
bea-lai2001, klo-bar-gre-one2013}.  We use a Barycentric quadrature
rule~\cite{bar2014, bar-wu-vee2015} since it is a non-intrusive
modification of the trapezoid rule, and the error is guaranteed to be
uniformly bounded.  We extend the original quadrature method so that it
can be used to compute the gradient of the velocity that is required to
compute the shear stress and the fluid vorticity.

\begin{figure}[H]
\begin{center}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b1}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b80}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b160}\\
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b240}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b320}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b400}
\caption{\label{fig:Eroding50vort} The same six snapshots as
Figure~\ref{fig:Eroding50vel}.  The color is the vorticity of the fluid.
Since the rate of erosion is equivalent to the magnitude of the
vorticity, erosion is fastest in the yellow and blue regions and slowest
in the green regions.}
\end{center}
\end{figure}

To characterize the transport, particle trajectories need to be
computed. Depending on the application, microscale transport can be
modelled as pure advection~\cite{dea-qua-bir-jua2018,
leb-ded-dav-bou2007, cve-che-wen1996, puy-gou-den2019},
advection-diffusion~\cite{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\cite{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume that the trajectories are governed purely by
advection (ie.~no diffusion). Therefore, the particle trajectories are
identical to the streamlines, and they can be found, for example, with a
continuous time random walk~\cite{mon-wei1965, den-cor-sch-ber2004,
leb-den-car2008, ber-cor-den-sch2006}.  Instead,  we compute
trajectories $\ss(t)$ that are initialized at $\ss_0$ by solving
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity.  Since there is no stiff diffusive
term in~\eqref{eqn:tracers}, we use a fourth-order explicit Runge-Kutta
time stepping method.  In addition, the Barycentric quadrature rule is used to
accurately compute trajectories that are close to an eroding grain.

Once streamlines in an eroded geometry are computed, we characterize the
transport by computing and analyzing three different metrics: the
tortuosity, the anomalous dispersion, and the pore size distribution.
The local tortuosity of a streamline connecting the inlet to the outlet
is the streamline's length normalized by the distance between the inlet
and the outlet.  In porous media, depending on several factors including
the porosity, the local tortuosity can be greater than
1.5~\cite{kop-kat-tim1996, mat-kha-koz2008}, or even
2~\cite{dud-koz-mat2011}.  The tortuosity of a geometry is defined by
averaging the local tortuosity over all streamlines that are initialized
at the inlet, and the geometry's tortuosity can be used to characterize
average particle motions~\cite{hak-com-den2019}.  To characterize
spreading, the fluid dispersion is computed by calculating the variance
of the streamline lengths. In porous media, this spreading can often be
characterized with a super-diffusive anomalous dispersion
rate~\cite{kan-dea-nun-bij-blu-jua2014, cus-hu-den1995,
dea-leb-den-tar-bol-dav2013}. An accurate solution near the grains is
crucial since the anomalous dispersion is caused by streamlines spending
time in both the high and low velocity regimes~\cite{ber-sch2001}.
Finally, the third metric, pore size distribution, is useful to quantify
velocity distributions~\cite{ali-par-wei-bre2017, dea-qua-bir-jua2018},
channelization~\cite{sie-ili-pri-riv-gua2019},
connectivity~\cite{knu-car2005, wes-blo-gra2001}, and to develop network
models~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
Our first contribution is extending the work of Barnett et
al.~\cite{bar-wu-vee2015} so that high-order accurate derivatives of
solutions of the Stokes equations can be computed.  Using this new
quadrature rule, we accurately compute erosion rates and stably simulate
erosion.  Moreover, we use the quadrature method to visualize the
vorticity in the fluid bulk.  The vorticity is convenient to visualize
since, in the zero Reynolds number limit, it is equivalent to the shear
stress on the eroding bodies, and therefore its magnitude is the erosion
rate.

Our second contribution is computing and analyzing trajectories in
eroded geometries.  These trajectories are used to characterize the
transport by computing the local tortuosity of trajectories, and the
tortuosity and dispersion rate of eroded geometries. By using the same
quadrature rule and a fourth-order Runge-Kutta method, we are able to
accurately compute trajectories that spend large amounts of time in the
slow velocity regions near the eroding bodies, and reliably computing
these trajectories is essential to characterize the
flow~\cite{dea-leb-den-tar-bol-dav2013}.  We compare our tortuosity and
dispersion rates with models available in the literature.  Finally, we
analyze the effect of erosion on the pore size distribution.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations}
As in our previous work~\cite{qua-moo2018}, we consider two-dimensional
immobile eroding grains. However, this work does not have a limitation
on the proximity of the grains. Compared to our previous work, we are
able to simulate much denser suspensions with a comparable number of
discretization points and without upsampling.  While we do investigate
transport in this work, we only consider passive particles that are
advected by the flow.  To model other physical processes, such as
contaminant transport, we plan to simulate a finite Peclet number
advection-diffusion equation in eroded geometries.
